Statistical Mechanics Informed Neural Networks for Solar Wind modeling

### 2024-12-09 paper/main.tex
hola

\subsection{Statistical-Physics-Informed Neural Networks (Stat-PINNs): A Machine Learning Strategy for Coarse-graining Dissipative Dynamics}

@@

\subsection{Boltzmann's Equation}

@@

\subsection{Vlasov's Equation}

@@

For this case particularly, the colisionless nature of the plasma and Liouville's theorem assumes no change in entropy, if we use Leibniz method and compute the time derivative of the integrant:

@@

@@

This models were created as numerical methods for solving Partial Differential Equations that embed physical modeling. It further extended to all fields that required PDE solving as a computationally efficient numerical alternative to other solvers.

@@

@@

\subsection{Loss function}
The neural network is trained using the standard gradient-descent based optimization algorithm: Adam. One Cycle learning rate scheduler imizer is used to facilitate the convergence to a learning rate plateau. L2 regularizer and dropout is implemented to avoid overfitting.

The loss function consists on a Liouville, probabilistic, and entropy term.

\begin{equation}
    \mathcal{L}_{Liouville}\left(\mathbf{E}, \mathbf{B}, \mathbf{v}, \mathbf{r}; \mathbf{\theta}\right) = \left(\frac{\partial f_{\alpha}}{\partial t} + \mathbf{v} \cdot \nabla_{\mathbf{r}}  f_{\alpha} + \frac{q (\mathbf{E} + \mathbf{v} \times \mathbf{B})}{m} \cdot \nabla_{\mathbf{v}} f_\alpha \right)^2
\end{equation}

\begin{equation}
    \mathcal{L}_{prob} \left(x; \mathbf{\theta}\right) = \left(\int_{x \sim X} f_\alpha (\mathbf{v}, \mathbf{r}, t) d^3\mathbf{v} d^3\mathbf{r} - 1\right)^2
\end{equation}

\begin{equation}
    \mathcal{L}_S(\mathbf{r}, \mathbf{v}, t) = \left(\frac{d}{dt}\int f_\alpha \log f_\alpha \right)^2
\end{equation}

\begin{equation}
    \mathcal{L}_{regularizer} \left( \mathbf{\theta} \right) = \mathbb{E}_{w \in \mathbf{\theta}}\left[ w ^ 2 \right]
\end{equation}

Getting our final objective function:

\begin{equation}
    \mathcal{L}\left[\mathcal{L}_{Liouville}, \mathcal{L}_{prob}, \mathcal{L}_{S} \mathcal{L}_{regularizer}\right] = \alpha_1 \mathcal{L}_{Liouville} + \alpha_2 \mathcal{L}_{prob} + \alpha_3 \mathcal{L}_{S} + \lambda \mathcal{L}_{regularizer}
\end{equation}

>>>>

@@

\subsection{Boltzmann-Vlasov Equations}

@@

@@

For this case particularly, the collisionless nature of the plasma and Liouville's theorem assumes no change in entropy, if we use Leibniz method and compute the time derivative of the integrant:

@@

\subsection{Maxwell's equations}

To complete the description of the solar wind, the Maxwell's electrodynamic formulation provides our final description to well-round the kinetic analysis of the plasma:

\begin{align*}
    \nabla \cdot E &= \frac{\sum_{\alpha} \rho_\alpha}{\epsilon_0} \\
    \nabla \cdot B &= 0 \\
    \nabla \times E &= - \frac{\partial B}{\partial t} \\
    \nabla \times B &= \mu_0 \left(J + \epsilon_0 \frac{\partial E}{\partial t}\right)
\end{align*}

\subsection{Complete dynamical system}
For some general universal approximator $f_n$, also interpreted as dynamical descriptor in this context, there is a parameter space $\theta_n$ tied to the governing constraints ($g_i$) for its respective phenomena:

\begin{align*}
    f_n(\mathbf{r}, \mathbf{v}, t; \theta_n): \nabla f_n = \sum_{i} \lambda_i \nabla g_i
\end{align*}

For this sake, we employ a Statistical mechanics informed Vlasov-Maxwell model that fully describes the known, ideal behavior of the solar wind:

\begin{align*}
    \frac{\partial f_\alpha}{\partial t} + f_v \cdot \nabla_{\mathbf{r}} f_\alpha &+ \frac{q_\alpha ( f_E + f_v \times f_B)}{m_\alpha} \cdot \nabla_{\mathbf{v}} f_\alpha = 0 \text{ (Vlasov)}\\
    \sum_{\alpha} q_\alpha \int &f_\alpha d^3(f_v) = 0 \text{ (Quasineutrality)}\\
\frac{d}{dt} \int &f_\alpha \log f_\alpha d^3r d^3(f_v) = 0 \text{ (Liouville)}\\
\nabla \cdot f_E &= \frac{\sum_{\alpha} \rho_\alpha}{\epsilon_0} \\
\nabla \cdot f_B &= 0 \\
\nabla \times f_E &= - \frac{\partial f_B}{\partial t} \\
\nabla \times f_B &= \mu_0 \left(J + \epsilon_0 \frac{\partial f_E}{\partial t}\right) \text{ (Maxwell)} \\
\end{align*}

$f_\alpha$ is the kinetic description of each specie, $f_E$ is the electric field descriptor, $f_B$ is the magnetic field descriptor, $f_v$ is the velocity descriptor, $J$ is the current density (derived from $f_\alpha$), and $\rho_\alpha$ is the charge density (derived from $f_\alpha$).

The cross-dependance between both models for the backpropagation process is crucial; thus, a backbone architecture for the electromagnetic descriptor engineered to exploit the backpropagation of the kinetic model can lead to faster convergence in addition to the Maxwell constraints.

A cross-training process that relies on the bonds between the Maxwell and Vlasov PDE criterions will be critical; the PDE loss APORTE to the training process will be gradually incremented through epochs, and inversely correlated with the distances from the boundary contour delimiting the training space to minimize the error margin of non-representative results in the middle of the training process.

This structure provides an optimization process overtly relying on PDE residuals; therefore, an appropiate set of boundary conditions is crucial for the convergence of the physics informed neural networks.

\subsubsection{Scope and boundary conditions}
The current work utilizes data from L1-Lagrange readings; accordingly, the phase space is constrained to fit predictions up to the Earth's position. As so, we define our parameter space:

\begin{align*}
    r &\in \left[10R_E, 240R_E\right] \\
    v_x &\in \left[300, 1000\right]\frac{km}{s} \\
    t &\in \left[0, 7200]s \\
\end{align*}

For the kinetic and electormagnetic model, we define the following Dirichlet boundary conditions:

\begin{align*}
    f_\alpha\left(r, v_{max}, t) &= 0 \\
    f_\alpha\left(r, v_{min}, t) &= 0 \\
    f_v\left(r_{min}, t) &= \mathbf{v}_{L1}(t) \\
    f_v\left(r_{max}, t) &= \mathbf{v}_{Earth}(t) \\
    f_E\left(r_{min}, t\right) &= E_{L1}(t) \\
    f_E\left(r_{max}, t\right) &= E_{Earth}(t) \\
    f_B\left(r_{min}, t\right) &= B_{L1}(t) \\
    f_B\left(r_{max}, t\right) &= B_{Earth}(t)
\end{align*}

@@

These models were created as numerical methods for solving Partial Differential Equations that embed physical modeling. It further extended to all fields that required PDE solving as a computationally efficient numerical alternative to other solvers.

@@

This model establishes a cornerstone for solar wind modeling, bonding the kinetic model distribution to the results of the electromagnetic model, extending the training space from     L1 Lagrange to a whole grid interval from L1 to the Earth.

@@
